MSstatsPTM: Statistical relative quantification of post-translational modifications in bottom-up mass spectrometry-based proteomics

### 2022-07-12 code/Spike-in_Analysis.Rmd
small updates to results

@@

ttest_final_model$trueLog2FC=ttest_final_model$trueLog2FC*-1
ttest_final_model[ttest_final_model$Label == 'mix1-mix2'][['trueLog2FC']] <- ttest_final_model[ttest_final_model$Label == 'mix1-mix2'][['trueLog2FC']]*-1
ttest_final_model[ttest_final_model$Label == 'mix1-mix3'][['trueLog2FC']] <- ttest_final_model[ttest_final_model$Label == 'mix1-mix3'][['trueLog2FC']]*-1

@@

>>>>

combo_plot_df
combo_plot_df[combo_plot_df$Label == 'mix2-mix3'][['trueLog2FC']] <- combo_plot_df[combo_plot_df$Label == 'mix2-mix3'][['trueLog2FC']]*-1

@@

ttest_final_model$trueLog2FC=ttest_final_model$trueLog2FC*-1
ttest_final_model[ttest_final_model$label == 'mix1-mix2'][['trueLog2FC']] <- ttest_final_model[ttest_final_model$label == 'mix1-mix2'][['trueLog2FC']]*-1
ttest_final_model[ttest_final_model$label == 'mix3-mix4'][['trueLog2FC']] <- ttest_final_model[ttest_final_model$label == 'mix3-mix4'][['trueLog2FC']]*-1
ttest_final_model[ttest_final_model$label == 'mix1-mix3'][['trueLog2FC']] <- ttest_final_model[ttest_final_model$label == 'mix1-mix3'][['trueLog2FC']]*-1

ttest_final_model$contrast_truth = factor(ttest_final_model$contrast_truth, 
                                          levels = 
                                  c("LogFC: -1 (mix2-mix1)",
                                    "LogFC: 1 (mix3-mix1)",
                                    "LogFC: 0 (mix4-mix1)",
                                    "LogFC: 2 (mix3-mix2)",
                                    "LogFC: 1 (mix4-mix2)",
                                    "LogFC: -1 (mix4-mix3)"))

@@

spike[spike$label == 'mix1-mix2'][['contrast_truth']] <- "LogFC: -1 (mix2-mix1)"
spike[spike$label == 'mix1-mix3'][['contrast_truth']] <- "LogFC: 1 (mix3-mix1)"
spike[spike$label == 'mix1-mix4'][['contrast_truth']] <- "LogFC: 0 (mix4-mix1)"
spike[spike$label == 'mix2-mix3'][['contrast_truth']] <- "LogFC: 2 (mix3-mix2)"
spike[spike$label == 'mix2-mix4'][['contrast_truth']] <- "LogFC: 1 (mix4-mix2)"
spike[spike$label == 'mix3-mix4'][['contrast_truth']] <- "LogFC: -1 (mix4-mix3)"

spike[spike$Label == 'mix3-mix4'][["Label"]] = "mix4 vs mix3"
spike[spike$Label == 'mix2-mix4'][["Label"]] = "mix4 vs mix2"
spike[spike$Label == 'mix1-mix4'][["Label"]] = "mix4 vs mix1"
spike[spike$Label == 'mix2-mix3'][["Label"]] = "mix3 vs mix2"
spike[spike$Label == 'mix1-mix3'][["Label"]] = "mix3 vs mix1"
spike[spike$Label == 'mix1-mix2'][["Label"]] = "mix2 vs mix1"

spike[spike$Label == "mix3 vs mix2" & 
        spike$Model == "T-test"][["trueLog2FC"]] = 2
spike[spike$Label == "mix3 vs mix2" & 
        spike$Model == "T-test"][["log2FC"]] = spike[
          spike$Label == "mix3 vs mix2" & 
        spike$Model == "T-test"][["log2FC"]] * -1
spike[spike$Label == "mix4 vs mix2" & 
        spike$Model == "T-test"][["trueLog2FC"]] = 1
spike[spike$Label == "mix4 vs mix2" & 
        spike$Model == "T-test"][["log2FC"]] = spike[
          spike$Label == "mix4 vs mix2" & 
        spike$Model == "T-test"][["log2FC"]] * -1
spike[spike$Label == "mix4 vs mix3" & 
        spike$Model == "T-test"][["trueLog2FC"]] = -1
spike[spike$Label == "mix4 vs mix3" & 
        spike$Model == "T-test"][["log2FC"]] = spike[
          spike$Label == "mix4 vs mix3" & 
        spike$Model == "T-test"][["log2FC"]] * -1

spike[spike$Label == "mix3 vs mix2" & 
        spike$Model == "Limma"][["trueLog2FC"]] = 2
spike[spike$Label == "mix4 vs mix2" & 
        spike$Model == "Limma"][["trueLog2FC"]] = 1
spike[spike$Label == "mix4 vs mix3" & 
        spike$Model == "Limma"][["trueLog2FC"]] = -1

### 2022-07-12 main/MSstatsPTM_Main.tex
small updates to results

Interestingly, the test above showed that the unmodified protein was more likely to cause observed changes in PTM abundance, rather than mask changes in the PTM. The later case occurs when the unmodified protein exhibits a change in abundance, while the modification does not. An example of this can be seen in \figref{data5_profile_plot}. Originally the modification at $S178$ of protein $TTP$ was shown to be differentially abundant between KO\_Uninfect and KO\_Early, with a log fold change of $2.9$. However the unmodified protein was shown to contribute $2.014$ of this change, meaning that the modification only accounted for $.886$ after adjustment. This caused the modification to go from differentially abundant to non-differentially abundant. As discussed previously, these cases occur frequently and can be thought of as false positives which are caused by confounding with the unmodified protein. The proposed approach corrects for this confounding, allowing us to see the true fold change of PTMs, while excluding those that do not exhibit differential changes.

Conversely, cases where the unmodified protein masks changes in the PTM are more infrequent. In this case, the modification and unmodified protein exhibit contradictory changes in abundance, moving in opposite directions. One such case can be seen in~\figref{data4_profile_plot}. Luchetti et al.~\cite{LUCHETTI2021} showed that $GSDMD$ was actively degraded when IpaH7.8 expression was induced by dox treatment. Our reanalysis confirmed this, with the $GSDMD$ protein being downregulated when Dox treatments reached the 4 and 6 hour marks. Conversely, ubiquitination of $GSDMD$ at site $K62$ upregulated abundance between the same conditions. This upregulation was originally confounded by the downregulation of unmodified $GSDMD$, appearing to have little change between no Dox and Dox 4 and 6 hour. The proposed approach accounted for this confounding and the modification was shown to be differentially abundant, with an absolute fold change of $1.415$ between the Dox 1 hour and Dox 4 hour conditions. This was an interesting result, as the modification contradicts what previous research showed for the unmodified protein. This effect can be thought of as a false negative and would have been challenging to observe without the proposed approach. There are many non-differentially abundant PTMs, and having to comb through each one individually to determine false negatives is generally an unfeasible task. The proposed approach removes the confounding on an experimental level, revealing interesting PTMs that are entirely masked due to changes in the unmodified protein.

>>>>

\figref{venn_diagrams} demonstrates that the unmodified protein was more likely to cause the observed changes in PTM abundance, rather than mask changes in the PTM. The later case can occur when the unmodified protein exhibits a change in abundance, while the modification does not. Additionally, this case may come about due to the high variability of PTMs compared to the global protein. An example of this can be seen in \figref{data5_profile_plot}. Originally the modification at $S178$ of protein $TTP$ was shown to be differentially abundant between KO\_Uninfect and KO\_Early, with a log fold change of $2.9$. However the unmodified protein was shown to contribute $2.014$ of this change, meaning that the modification only accounted for $.886$ after adjustment. This caused the modification to go from differentially abundant to non-differentially abundant. These cases occur frequently and can be thought of as false positives. The proposed approach corrects for this confounding, allowing us to see the true fold change of PTMs, while excluding those that do not exhibit differential changes.

Conversely, cases where the unmodified protein masks changes in the PTM are more infrequent. In this case, the modification and unmodified protein exhibit contradictory changes in abundance, moving in opposite directions. One such case can be seen in~\figref{data4_profile_plot}. Luchetti et al.~\cite{LUCHETTI2021} showed that $GSDMD$ was actively degraded when IpaH7.8 expression was induced by dox treatment. Our reanalysis confirmed this, with the $GSDMD$ protein being down-regulated when Dox treatments reached the 4 and 6 hour marks. Conversely, ubiquitination of $GSDMD$ at site $K62$ up-regulated abundance between the same conditions. This up-regulation was originally confounded by the down-regulation of unmodified $GSDMD$, and made the modification appear to have little change between no Dox and Dox 4 and 6 hour conditions. The proposed approach accounted for this confounding and the modification was shown to be differentially abundant, with an absolute fold change of $1.415$ between the Dox 1 hour and Dox 4 hour conditions. This was an interesting result, as the modification contradicts what previous research showed for the unmodified protein. This effect can be thought of as a false negative and would have been challenging to observe without the proposed approach. There are many non-differentially abundant PTMs, and having to comb through each one individually to determine false negatives is generally an unfeasible task. The proposed approach removes the confounding on an experimental level, revealing interesting PTMs that are entirely masked due to changes in the unmodified protein.
